\textbf{COMPARISON OF \\ HEURISTIC ALGORITHMS \\ FOR KP01} \\[0.5cm] \rule{12cm}{0.3mm} \\[0.5cm] \small \scshape{A comparison of heuristic algorithms for solving the 0-1 Knapsack Problem} \\[0.4cm] \rule{12cm}{0.3mm} \vskip -0.3cm

### 2023-04-15 doc/paper.tex
finished both algs in doc

% time complexities graph
\begin{center}
    \begin{tikzpicture}
      \begin{axis}[
          grid = major,
          clip = true,
          ticks = none,
          width=1\textwidth,
          height=0.8\textwidth,
          every axis plot/.append style={very thick},
          axis line style = ultra thick,
          clip mode=individual,
          restrict y to domain=0:10,
          restrict x to domain=0:10,
          axis x line = left,
          axis y line = left,
          domain = 0.00:10,
          xmin = 0,
          xmax = 11,
          ymin = 0,
          ymax = 11,
          xlabel = n,
          ylabel = Computer Operations,
          xlabel style = {at={(axis description cs:0.5,-0.1)},anchor=south},
          ylabel style = {at={(axis description cs:-0.08,0.5)},anchor=north},
          label style = {font=\LARGE\bf},
          ]
          \addplot [
          samples=100, 
          color=green,
          ]
          {x}node[above,pos=1,style={font=\Large}]{$\mathcal{O}(n)$};
          \addplot [
          samples=100, 
          color=blue,
          ]
          {log2 x}node[above,pos=1,style={font=\Large}]{$\mathcal{O}(\log{}n)$};
          \addplot [
          samples=100, 
          color=red,
          ]
          {3*2^x}node[above,pos=1,style={font=\Large}]{$\mathcal{O}(3\cdot2^n)$};
          \addplot [
          samples=100, 
          color=violet,
          ]
          {1}node[above,pos=1,style={font=\Large}]{$\mathcal{O}(1)$};  
          \addplot [
          samples=100,
          color=orange, 
          ] gnuplot{gamma(x+1)} node[above,pos=1,style={font=\Large}]{$\mathcal{O}(n!)$};  
      \end{axis}
    \end{tikzpicture}
\end{center}

@@

\begin{breakablealgorithm}
\caption{The DGHS algorithm}\label{DGHS}
    \begin{algorithmic}[1]
        \State Set the harmony memory size $HMS$, the number of iterations $ITERATIONS$, and the minimum and maximum values of parameters $PAR$ and $HMCR$.
        \State Initialize the HM, and use \ref{harmonyRepair} to the generated harmonies. Calculate the totalProfit and totalWeight values for each harmony in HM.
        \State $iterator \gets 1$
        \While {$iterator <= ITERATIONS$}
            \State Record indices of the best and the worst harmonies in HM.
            \State Calculate parameters HMCR and PAR for the current iteration.
            \State Perform Algorithm \ref{harmonyGen} to produce a new harmony $\Vec{x}_{new}$
            \State Perform Algorithm \ref{harmonyRepair} to repair the new harmony $\Vec{x}_{new}$
            \If {$\Vec{x}_{new} \text{ is better than or equal to } \Vec{x}_{best}$}
                \State Replace $\Vec{x}_{best} \text{ with } \Vec{x}_{new}$
            \ElsIf {$\Vec{x}_{new} \text{ is better than or equal to } \Vec{x}_{worst}$}
                \State Replace $\Vec{x}_{worst} \text{ with } \Vec{x}_{new}$
            \EndIf
            \State $iterator \gets iterator+1$
        \EndWhile
        \State Output profit of best harmony
    \end{algorithmic}
\end{breakablealgorithm}

@@

\begin{breakablealgorithm}
\caption{Generating a new harmony during iterative part (part 2)}\label{harmonyGen}
    \begin{algorithmic}[1]
        \For {$i = 1 \text{ to } noItems$}
            \If {$rand(0, 1) <= HMCR(t)$}
                \State $newHarmony[i] \gets bestHarmony[i]$ 
            \Else
                \State Generate a random integer number $a \in \{1, 2, ... HMS\}, a \neq best$
                \State $newHarmony[i] \gets Harmony_{a}[i]$ 
                \If {$rand(0, 1) <= PAR(t)$}
                    \State $newHarmony[i] = |newHarmony[i]-1|$ \Comment{// Flipping i'th item - mutation}
                \EndIf
            \EndIf
        \EndFor
    \end{algorithmic}
\end{breakablealgorithm}

@@

\begin{breakablealgorithm}
\caption{Repair-operator after harmony generation}\label{harmonyRepair}
    \begin{algorithmic}[1]
        \If {$totalWeight > maxWeight$}
            \For {$i = 1 \text{ to } N$} \Comment{ DROP phase}
                \State $\lambda_{i} = \frac{profit_{i}}{weight_{i}}$
            \EndFor
            \State Sort items in increasing order of $\lambda_{i}$, and let $ind_{i}$ denote the original index of each $\lambda_{i}$
            \For {$i = 1 \text{ to } noItems $} 
            \State Remove the ones with the least profit-density values greedily
                \If {$\lambda_{i} == 0$}
                    \State \textbf{Continue}
                \EndIf
                \State $newHarmony[ind_{i}] = 0$ \Comment{Unload the item}
                \State $totalWeight \gets totalWeight - weight[ind_{i}]$
                \State $totalProfit \gets totalProfit - profit[ind_{i}]$
                \If {$totalWeight <= maxWeight$}
                    \State \textbf{Break} \Comment{Terminate DROP phase}
                \EndIf
            \EndFor
        \EndIf
        \If {$totalWeight < maxWeight$}
            \For {$i = 1 \text{ to } N$} \Comment{ADD phase}
                \State $\lambda_{i} = \frac{profit_{i}}{weight_{i}}$
            \EndFor
            \State Sort items in increasing order of $\lambda_{i}$, and let $ind_{i}$ denote the original index of each $\lambda_{i}$
            \For {$i = 1 \text{ to } noItems $} 
                \State Add the ones with the greatest profit-density if possible 
                \If {$newHarmony[ind_{i}] == 0$}
                    \If {$totalWeight + weight[ind_{i}] <= maxWeight$}
                        \State $newHarmony[ind_{i}] = 1$
                        \State $totalWeight \gets totalWeight + weight[ind_{i}]$
                        \State $totalProfit \gets totalProfit + profit[ind_{i}]$
                    \EndIf
                \EndIf
            \EndFor
        \EndIf
    \end{algorithmic}
\end{breakablealgorithm}

@@

>>>>

% time complexities graph
\begin{center}
    \begin{tikzpicture}
      \begin{axis}[
          grid = major,
          clip = true,
          ticks = none,
          width=1\textwidth,
          height=0.8\textwidth,
          every axis plot/.append style={very thick},
          axis line style = ultra thick,
          clip mode=individual,
          restrict y to domain=0:10,
          restrict x to domain=0:10,
          axis x line = left,
          axis y line = left,
          domain = 0.00:10,
          xmin = 0,
          xmax = 11,
          ymin = 0,
          ymax = 11,
          xlabel = n,
          ylabel = Computer Operations,
          xlabel style = {at={(axis description cs:0.5,-0.1)},anchor=south},
          ylabel style = {at={(axis description cs:-0.08,0.5)},anchor=north},
          label style = {font=\LARGE\bf},
          ]
          \addplot [
            samples=100, 
            color=red,
          ]
          {3*2^x}node[above,pos=1,style={font=\Large}]{$\mathcal{O}(3\cdot2^n)$};
          \addplot [
            samples=100,
            color=orange, 
          ] gnuplot{gamma(x+1)} node[above,pos=1,style={font=\Large}]{$\mathcal{O}(n!)$};
          \addplot [
            samples=100, 
            color=green,
          ]
          {x}node[above,pos=1,style={font=\Large}]{$\mathcal{O}(n)$};
          \addplot [
            samples=100, 
            color=blue,
          ]
          {log2 x}node[above,pos=1,style={font=\Large}]{$\mathcal{O}(\log{}n)$};
          \addplot [
            samples=100,
            color=violet,
          ]
          {2*sqrt x}node[above,pos=1,style={font=\Large}]{$\mathcal{O}(2\sqrt{n})$};
          \addplot [
          samples=100, 
          color=red,
          ]
          {1}node[above,pos=1,style={font=\Large}]{$\mathcal{O}(1)$};

      \end{axis}
    \end{tikzpicture}
\end{center}

@@

%DGHS
\begin{breakablealgorithm}
\caption{The DGHS algorithm}\label{DGHS}
    \begin{algorithmic}[1]
        \State Set the harmony memory size $HMS$, the number of iterations $ITERATIONS$, and the minimum and maximum values of parameters $PAR$ and $HMCR$.
        \State Initialize the HM through a randomized process, and use \ref{harmonyRepair} to the generated harmonies. Calculate the totalProfit and totalWeight values for each harmony in HM.
        \State $iterator \gets 1$
        \While {$iterator <= ITERATIONS$}
            \State Record indices of the best and the worst harmonies in HM.
            \State Calculate parameters HMCR and PAR for the current iteration.
            \State Perform Algorithm \ref{harmonyGen} to produce a new harmony $\Vec{x}_{new}$
            \State Perform Algorithm \ref{harmonyRepair} to repair the new harmony $\Vec{x}_{new}$
            \If {$\Vec{x}_{new} \text{ is better than or equal to } \Vec{x}_{best}$}
                \State Replace $\Vec{x}_{best} \text{ with } \Vec{x}_{new}$
            \ElsIf {$\Vec{x}_{new} \text{ is better than or equal to } \Vec{x}_{worst}$}
                \State Replace $\Vec{x}_{worst} \text{ with } \Vec{x}_{new}$
            \EndIf
            \State $iterator \gets iterator+1$
        \EndWhile
        \State Output profit of best harmony
    \end{algorithmic}
\end{breakablealgorithm}

@@

%harmonyGen
\begin{breakablealgorithm}
\caption{Generating a new harmony during iterative part (part 2)}\label{harmonyGen}
    \begin{algorithmic}[1]
        \For {$i = 1 \text{ to } noItems$}
            \If {$rand(0, 1) <= HMCR(t)$}
                \State $newHarmony[i] \gets bestHarmony[i]$ 
            \Else
                \State Generate a random integer number $a \in \{1, 2, ... HMS\}, a \neq best$
                \State $newHarmony[i] \gets Harmony_{a}[i]$ 
                \If {$rand(0, 1) <= PAR(t)$}
                    \State $newHarmony[i] = |newHarmony[i]-1|$ \Comment{// Flipping i'th item - mutation}
                \EndIf
            \EndIf
        \EndFor
    \end{algorithmic}
\end{breakablealgorithm}

@@

%harmonyRepair
\begin{breakablealgorithm}
\caption{Repair-operator for DGHS}\label{harmonyRepair}
    \begin{algorithmic}[1]
        \If {$totalWeight > maxWeight$}
            \For {$i = 1 \text{ to } N$} \Comment{ DROP phase}
                \State $\lambda_{i} = \frac{profit_{i}}{weight_{i}}$
            \EndFor
            \State Sort items in increasing order of $\lambda_{i}$, and let $ind_{i}$ denote the original index of each $\lambda_{i}$
            \For {$i = 1 \text{ to } noItems $} 
            \State Remove the ones with the least profit-density values greedily
                \If {$\lambda_{i} == 0$}
                    \State \textbf{Continue}
                \EndIf
                \State $newHarmony[ind_{i}] = 0$ \Comment{Unload the item}
                \State $totalWeight \gets totalWeight - weight[ind_{i}]$
                \State $totalProfit \gets totalProfit - profit[ind_{i}]$
                \If {$totalWeight <= maxWeight$}
                    \State \textbf{Break} \Comment{Terminate DROP phase}
                \EndIf
            \EndFor
        \EndIf
        \If {$totalWeight < maxWeight$}
            \For {$i = 1 \text{ to } N$} \Comment{ADD phase}
                \State $\lambda_{i} = \frac{profit_{i}}{weight_{i}}$
            \EndFor
            \State Sort items in increasing order of $\lambda_{i}$, and let $ind_{i}$ denote the original index of each $\lambda_{i}$
            \For {$i = 1 \text{ to } noItems $} 
                \State Add the ones with the greatest profit-density if possible 
                \If {$newHarmony[ind_{i}] == 0$}
                    \If {$totalWeight + weight[ind_{i}] <= maxWeight$}
                        \State $newHarmony[ind_{i}] = 1$
                        \State $totalWeight \gets totalWeight + weight[ind_{i}]$
                        \State $totalProfit \gets totalProfit + profit[ind_{i}]$
                    \EndIf
                \EndIf
            \EndFor
        \EndIf
    \end{algorithmic}
\end{breakablealgorithm}

@@

%BMQHOA
\begin{breakablealgorithm}
\caption{The BMQHOA algorithm with solution generation}\label{BMQHOA}
    \begin{algorithmic}[1]
        \State Set the number of iterations $ITERATIONS$ and the number of binary vectors (solutions) - BINVEC in memory.
        \State Randomly generate the binary vectors and use Algorithm \ref{harmonicRepair} to repair the vectors. 
        \While {$iterator <= ITERATIONS$}
            \State Update $\sigma_{s}$
            \State $found \gets FALSE$
            \While {$found == FALSE$}
                \State Try to generate a solution
                \State Let $solutions_{new}$ be the new generated vector 
                \State $solutions_{new} \gets solutions_{best}$
                \State Generate the number of flipped bits $m ~ N(0, \sigma_{s})$
                \State Treat $solutions_{new}$ as a circular array
                \State Randomly select a position in $solutions_{new}$ and flip the next $m$ items.
                \State Mutate a random bit towards current best solution (flip an item)
                \State $solutions_{new}[rand] \gets solutions_{best}[rand]$
                \State Repair newly generated solution
                \If {$totalProfit >= totalProfit_{worst}$}
                    \State Replace worst solution with newly generated solution
                    \State $solutions_{worst} \gets solutions_{new}$
                    \State $found = TRUE$
                \EndIf
            \EndWhile
        \EndWhile
    \end{algorithmic}
\end{breakablealgorithm}

After generating a new solution, $m$ bits are flipped in the hope of increasing diversity in the solutions and increasing the likelihood of finding the optimal solution. They are generated randomly with a normal probability distribution around 0, with a variable std deviation. The Std. Deviation decreases with each new-solution-generation iteration so as to increase diversity in the beginning of the search, while reducing the likelihood of corrupting a binary vector closer to the end of the search when close to optimal solutions have (hopefully) been found. The normal distribution probability density function generator is implemented with the C++ Standard Library class normal\_distribution. An item's decision value is also made to match the corresponding decision in the best solution, which gives a slow mutation towards the current best solution allowing for diversity while still allowing the best solution to influence the process. Allowing multiple bits to mutate toward current best, this can lead to a premature local maximum, nullifying the rest of the search \ref{BMQHOA-article} .

The repair-operator of the BMQHOA algorithm has 3 phases:
\begin{enumerate}
    \item Density-first stage: The already selected items are sorted based on their profit to weight ratio in non-increasing order and then greedily selected while respecting the weight constraint. 
    \item Minimum-weight-first stage: Out of the items that weren't selected in the first stage are sorted based on their weight values in non-decreasing order, then greedily selected while respecting the weight constraint.   
\end{enumerate}

$Q_{1}[1, 2,...,n]$ is the index for the items in the density ratio array in the original vector. $Q_{2}[1, 2,..., n]$ is index of the items in the minimum weight sorted array in the original vector.

\begin{breakablealgorithm}
\caption{Repair-Operator for BMQHOA} 
    \begin{algorithmic}
        \State Let $x$ be the current array/vector.
        \State $totalWeight = 0, temp = 0, i = 0, b = 0$
        \State Stage 1: Density first stage
        \While {$temp < maxWeight$}
            \State $totalWeight = temp$
            \State $i += 1$
            \State $temp = temp + weight[Q_{1}[i]]$
        \EndWhile
        \For {$j = i \text{ to } n$}
            \State $x[Q_{1}[j]] = 0$
            \State $totalWeight += x[Q_{1}[j]]$
        \EndFor
        \State Stage 2: Minimum weight first stage
        \While {$totalWeight < C$}
            \State $b += 1$
            \If {$totalWeight + weight[Q_{2}[b]] <= C$} 
                    \State $x[Q_{2}[b]] = 1$
                \State $totalWeight += weight[Q_{2}[b]]$
            \EndIf
        \EndWhile
    \end{algorithmic}
\end{breakablealgorithm}
